%Abstract will go here.  Make sure it remains within the 350 word limit.
%\par To get a new paragraph
The ATLAS experiment at the Large Hadron Collider has entered the precision era of Higgs boson physics. Run 3 data collection, combined with the upcoming High-Luminosity LHC, places new demands on both the physics analyses and the reconstruction algorithms. Advances in heavy-flavor jet tagging using machine learning have directly improved the sensitivity of Higgs coupling measurements.
\par
Graph neural network and Transformer-based flavor taggers (GN1, GN2, GN2X) represent a significant advance over classical multivariate methods, leveraging low-level track information to substantially outperform previous approaches. The $t\bar{t}$ mistag rate calibration of GN2X, performed as the author's ATLAS qualification task, yielded scale factors consistent with unity across all probe-jet $p_\mathrm{T}$ bins, enabling deployment in Run 2 and Run 3 analyses.
\par
These taggers directly enable two hadronic Higgs analyses. In the $VH$ channel, the addition of a new low-$p_\mathrm{T}$ signal region contributed to the first ATLAS observation of $Z \rightarrow c\bar{c}$ at 5.2 $\sigma$ and set an observed upper limit of $|\kappa_c| < 4.2$ on the Higgs charm Yukawa coupling. In the vector boson fusion channel, an adversarial neural network classifier enabled the first evidence for $\mathrm{VBF}\,H \rightarrow b\bar{b}$ at 3.2 $\sigma$ in the VBF production mode.
\par
At $\langle\mu\rangle = 200$, traditional pileup mitigation algorithms degrade severely. Two event-wide architectures are introduced: PUMiNet uses attention to directly provide jet energy and mass correction factors, and PhyGHT extends this with physics-informed hypergraph attention to outperform all baselines, recovering the Higgs boson, $W$ boson, and top quark mass peaks under HL-LHC pileup conditions.
\par
Finally, a Transformer regression model trained on unpolarized $t\bar{t}$ is shown to extract a measurable left/right asymmetry in $\cos\theta$ for both the bottom and down quarks in hadronic top decays, demonstrating that spin information can be recovered without polarization-dependent training data and providing a path toward hadronic top polarimetry in full ATLAS simulation.
