I am deeply grateful for the opportunity to pursue a PhD and contribute to the largest data science experiment in human history during a data science revolution. It is a rare and fortunate collision of circumstances.

My advisor, Dr.~Flera Rizatdinova, gave me the freedom to pursue ideas at the boundary of high energy physics and computer science, and the support to see them through. I also thank my committee members, Dr.~Alexander Khanov, Dr.~Dorival Gonçalves, and Dr.~Atriya Sen, for their time and insight, and Mohammed Rakib for a rewarding collaboration at the intersection of high energy physics and machine learning.

Experimental physics has only begun to tap into the expressive power of modern machine learning, and I believe the most impactful applications are still ahead. The HL-LHC will demand every tool we are developing today, and I feel fortunate to be entering the field at exactly this moment.

Finally, I would like to thank my family for supporting me throughout my education. Very few students from Oklahoma get to see the world through the lens of international science. I traveled to CERN, worked alongside physicists from dozens of countries, and spent years thinking about collisions happening 100 meters underground in Switzerland. I am deeply grateful that this PhD gave me that chance.

Portions of this dissertation were refined with the assistance of AI writing tools. All scientific work, analysis, and intellectual contributions are original and solely those of the author.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% NOTICE %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% If you Acknowledgments span MORE than one page, the graduate college
% requires that you have this footer present on every page. In order to put it
% there, you will put this footer command in some of the text on page 2 (or 3,
% etc.) in order to ensure a copy of the footer appears on each page:

%\blfootnote{Acknowledgments reflect the views of the author and are not endorsed
% by committee members or Oklahoma State University.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
